---
title: "P2P sandybeach: Data read and transformation"
author: 'E. Klein'
date: "March 2019"
output: 
  html_document:
    toc:  TRUE
    toc_float: TRUE
    theme: united
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
require(knitr)
options(width = 100, stringAsFactors=F)
opts_chunk$set(echo =T, message = F, error = F, warning = F, comment = NA,  
               fig.align = 'left',  fig.width = 7.5, fig.height = 6,
               tidy = F, cache.path = '.cache/', fig.path = 'fig/')
               
library(RColorBrewer)
palette(brewer.pal(8, "Set2"))

library(readxl)
library(dplyr)
library(lubridate)
library(reshape2)
library(leaflet)
library(readr)



## function to remove all spaces in a string
NoSpaces = function(x){
  return(gsub(" ", "", x))
}
```

Last run `r lubridate::now()`


```{r setfilename}

## add here the names of the file to analyse, with the correct path

## add the file name here, with eh correct path
## rocky or beach
baseDataDir =  "../data2"
baseEcosystem = "beach"
fileName = "URU-BCHUY.xlsx"

datafileName = file.path(baseDataDir, baseEcosystem, fileName)

## taxa dictionary file name
taxaDictFileName = file.path(baseDataDir, "taxalist", baseEcosystem, "SB-taxonlistALL_2.csv")

taxaDict = read_tsv(taxaDictFileName)

```

Read the site information

```{r readsiteinfo}

## Reading data 
beachData = read_xlsx(path = datafileName, sheet = 1)

## standardise the variable types and format
beachData$Country = toupper(beachData$Country)
beachData$City = toupper(beachData$City)
beachData$Beach = toupper(beachData$Beach)
beachData$Zone = toupper(beachData$Zone)

## remove the parenthesis fgrom the variable name
names(beachData)[20] = "Level"

siteDF = beachData[,1:21]
nRecords = nrow(beachData)

## create eventID: countryCode-LocalityCode-SiteCode-strataCode-yymmdd
eventID = with(beachData, paste(NoSpaces(Country), NoSpaces(City), NoSpaces(Beach), Station, 
                                sprintf("T%02i", Transect), sprintf("L%02i",Level), Zone,
                                paste0(year(Date), sprintf("%02i", month(Date)), sprintf("%02i", day(Date))), 
                       sep = "-"))


## put eventID as first variable
beachData = cbind(eventID, beachData)
siteDF = cbind(eventID, siteDF)
```

### Map of the Sites


Map of the sites

```{r sitesMap}
siteDF$Latitude = as.numeric(siteDF$Latitude)
siteDF$Longitude = as.numeric(siteDF$Longitude)

siteCoords = siteDF %>% dplyr::group_by(City, Beach, Zone) %>% 
  dplyr::summarise(lng = mean(Longitude, na.rm=T), 
            lat = mean(Latitude, na.rm=T))
leaflet(siteCoords) %>% addTiles() %>% addMarkers(label = ~paste0(Beach, "-", Zone))

```




## Read the Abundance data sheet

You need to check the taxonomy of your species names before thos step. To need to provide a file with the taxon match output from WoRMS. YOu only change the original taxon name if you cannot find it in Worms at the match. For all others cases where you can resolve the name using the online tools, you have to keep the original name in the excel table.

```{r readabundance}

## lets transform the DF into a long format using reshape
Occurrence = melt(beachData, id.vars = c(1,3:5,11,12,16,19:22), measure.vars = 23:ncol(beachData), 
                    variable.name = "scientificName", value.name = "abundance", na.rm = T)

## remove ecords with abundance==0
## uncomment if wants to remove abundance==0
## Occurrence= Occurrence %>% dplyr::filter(abundance!=0)

```


### Add the taxon name info from the WoRMS matched file

You have to have the file with the original scientific names and the accepted scientific name and the LSID from WoRMS. Please refer to the P2P web site to refresh how to do that.

The name of the matched taxa file is specified in the file names chunk

```{r readtaxonmatched}

## join the taxon fields to the abundance data
Occurrence = left_join(Occurrence, taxaDict, by = c("scientificName"="ScientificName"))

```

### Create eventID and occurrenceID

`eventId` is the combination of `Country`, `locality`, `site`, `Station`, `Transect` and `Level`.
For the `occurrenceID` is the same `eventID` but with a serial number of the records added

```{r createIDs}


## create occurrenceID, adding the seq number
organismSeq = 1:nrow(Occurrence)
occurrenceID = paste(eventID, sprintf("%000005i", organismSeq), sep="-")

Occurrence = cbind(occurrenceID, Occurrence)


```

### save file for analysis

```{r savefiles}
fileNameBeach = file.path(baseDataDir, "DataAnalysisFiles", baseEcosystem, 
                          paste(Occurrence$Country[1], NoSpaces(Occurrence$City[1]), NoSpaces(Occurrence$Beach[1]),
                                sep="-"))


readr::write_csv(path = paste0(fileNameBeach, "_beach_siteDF.csv"), siteDF)
readr::write_csv(path = paste0(fileNameBeach, "_beach_occurrence.csv"), Occurrence)

```


## Create DwC EVENT, OCCURRENCE and eMoF files. TO BE CODED

We will use abundance and cover in the eMoF extension. The vocabularies from OBIS [BODC NERC](http://vocab.nerc.ac.uk/collection/Q01/current/) corresponding to the methods, instruments and units are:

1. quadrat: 
2. abundance:
3. cover:

### Event core file

the minimum set of mandatory fields for the event core are:

- eventID	
- eventDate	
- samplingProtocol: from oceanbestpractices: SARCE http://dx.doi.org/10.25607/OBP-5
- samplingSizeValue: 0.25 
- samplingSizeUnit: square meters, from BODC vocabs.

```{r}
## create Grandparent eventID: SITE
  eventFile = siteDF %>% group_by(eventID) %>% 
    summarise(parentEventID = "", 
              eventDate = unique(Date),
              habitat = "Sandy Beach",
              country=unique(Country),
              locality = unique(City),
              site = unique(Beach),
              strata = unique(Zone),
              decimalLongitude = mean(Longitude),
              decimalLatitude = mean(Latitude),
              coordinateUncertaintyInMeters = "",
              geodeticDatum = "WGS84")
readr::write_csv(path = file.path(baseDataDir, "IPTFiles/beach", paste0(siteDF$Country[1], "-", siteDF$City[1], "-" , siteDF$Beach[1],"_ipt_event.csv")), eventFile)

```


```{r}
## create occurrence file
### Occurrence extension file
occurrenceFile = data.frame(occurrenceID = Occurrence$occurrenceID, 
                            eventID = Occurrence$eventID,
                            scientificName = Occurrence$ScientificName_accepted,
                            scientificNameID = Occurrence$LSID,
                            basisOfRecord = rep("HumanObservation", nrow(Occurrence)),
                            occurrenceStatus = rep("present", nrow(Occurrence)))
readr::write_csv(path = file.path(baseDataDir, "IPTFiles/beach", paste0(siteDF$Country[1], "-", siteDF$City[1],"-" , siteDF$Beach[1], "_ipt_occurrence.csv")), occurrenceFile)

```


```{r}
## create eMoF file for abundance
MoFFile = data.frame(occurrenceID = Occurrence$occurrenceID, 
                         eventID = Occurrence$eventID,
                         measurementType = rep("abundance", nrow(Occurrence)),
                         measurementTypeID = rep("http://vocab.nerc.ac.uk/collection/P06/current/UMSQ/1/",
                                                 nrow(Occurrence)),
                         measurementValue = as.numeric(Occurrence$abundance), 
                         measurementUnit = rep("count", nrow(Occurrence)),
                         measurementUnitID = rep("count", nrow(Occurrence))) ## needs to be checked
readr::write_csv(path = file.path(baseDataDir, "IPTFiles/beach", paste0(siteDF$Country[1], "-", siteDF$City[1],"-" , siteDF$Beach[1], "_ipt_MoF.csv")), MoFFile)

```

